---
title: "Global State Bandit Simulation"
author: "Marvin Ernst, Oriol Gelabert, Melisa Vadenja"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
library(coda)
library(tidyverse)
library(here)
library(glue)
library(future.apply)
library(stats)
library(nimble)
plan(multisession)
```

```{r}
output_dir <- "plots"

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}
```


# Problem Setup: Global Latent State

We simulate a $K$-armed bandit problem where **all arms share the same underlying latent state process** governed by a Hidden Markov Model (HMM). This represents a scenario where there's a common environment or external condition that affects the reward probability of all arms.

### Loading all the Models and Functions
```{r}
source(here("src/load_all.R"))
```

Setting a seed:
```{r}
set.seed(123)
```

---

# GLOABL SETTING

Number of arms:
```{r}
K <- 2
```

Number of time steps:
```{r}
N <- 5000
```

Transition Probability:
```{r}
pi_global <- matrix(c(0.999, 0.001,
                     0.001, 0.999),
                     nrow = 2, byrow = TRUE)  
```

Probability of Reward:
```{r}
mu <- matrix(c(0.95, 0.1,
               0.1, 0.95),
                nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Global State

## 1.1 Simulate Ground Truth

For only one Run:
```{r simulate-global-state}
generate_global_datasets(
  K = K,
  N = N,
  mu = mu,
  pi_global = pi_global,
  n_runs = 1,
  scenario_name = "single_run",
  root_path = "data_global"
)
```

Load the generated data:
```{r}
truth <- readRDS("data_global/single_run/global_truth_1.rds")
y_global <- truth$y
z_global <- truth$z
```


# 1.2 Visualize Latent States

```{r visualize-state}
qplot(1:N, z_global, geom = "line") +
  labs(title = "Latent Global State Over Time",
       x = "Time", y = "State")
```

## 1.3 Visualize Rewards Per Arm

```{r visualize-rewards}
y_df <- as.data.frame(y_global)
y_df <- y_df |> mutate(arm = factor(1:K)) |> 
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "reward") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(y_df, aes(x = time, y = reward, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Reward Streams Per Arm",
       x = "Time", y = "Reward")
```

---

# 2. Single Run - Model Comparison

## 2.1 Thompson Sampling

### Baseline
```{r}
baseline_ts_results <- bandit_baselines("ts", 
                                        K, N, y_global, z_global, mu, 
                                        dynamics = "common",
                                        batch_size = 1)
```

Store in df:
```{r}
baseline_ts_results_df <- data.frame(
  time = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward = baseline_ts_results$cumulative_reward,
  cumulative_regret = baseline_ts_results$cumulative_regret,
  model = "Baseline TS"
)
```

### Adavanced Model
```{r}
system.time({
res_adv_ts_common <- thompson_advanced(K, N, mu, y_global, z_global, 
                                       batch_size = 100, 
                                       burn = 1000, n_iter = 200,
                                       dynamics = "common")
                                      # window_size = 3100)
})
```

Store in df:
```{r}
adv_ts_results_df <- data.frame(
  time = seq_along(res_adv_ts_common$cumulative_reward),
  cumulative_reward = res_adv_ts_common$cumulative_reward,
  cumulative_regret = res_adv_ts_common$cumulative_regret,
  model = "Advanced TS"
)
```

Combining results into one df:
```{r}
ts_compare_df_s1 <- bind_rows(baseline_ts_results_df, adv_ts_results_df)
```

### Cumulative Reward:
```{r}
plotts<-ggplot(ts_compare_df_s1, aes(x = time, y = cumulative_reward, color = model)) +
  geom_line(linewidth = 1) +
  labs(title = "Cumulative Reward: Baseline vs Advanced Thompson Sampling",
       x = "Time",
       y = "Cumulative Reward") +
  theme_minimal()

ggsave(file.path(output_dir, "ts_cumulative_reward_single.png"), plot = plotts, width = 8, height = 6, dpi = 300)
```

### Cumulative Regret:
```{r}
plotts<-ggplot(ts_compare_df_s1, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs Advanced Thompson Sampling",
       x = "Time",
       y = "Cumulative Regret") +
  theme_minimal()

ggsave(file.path(output_dir, "ts_cumulative_regret_single.png"), plot = plotts, width = 8, height = 6, dpi = 300)
```

```{r}
ts_compare_df_s1 <- ts_compare_df_s1 %>%
  group_by(model) %>%
  arrange(time) %>%
  mutate(inst_regret= c(0, diff(cumulative_regret)))

ggplot(ts_compare_df_s1, aes(x = time, y = inst_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Instantaneous Regret: Baseline vs Advanced Thompson Sampling",
       x = "Time",
       y = "Instantaneous Regret") +
  theme_minimal()
```
```{r}
inst_regret_diff_df <- ts_compare_df_s1 %>%
  select(time, model, inst_regret) %>%
  pivot_wider(names_from = model, values_from = inst_regret) %>%
  mutate(regret_diff = `Advanced TS` - `Baseline TS`)

# Plot the difference in instantaneous regret
ggplot(inst_regret_diff_df, aes(x = time, y = regret_diff)) +
  geom_line(color = "darkred", size = 1) +
  labs(title = "Difference in Instantaneous Regret (Advanced - Baseline)",
       x = "Time",
       y = "Regret Difference") +
  theme_minimal()
```

## 2.2 UCB

### Baseline
```{r}
baseline_ucb_results <- bandit_baselines(
  algorithm='ucb-tuned',
  K, N, y_global, z_global, mu, 
  dynamics = "common",
  batch_size = 1
  )
```

Store in df:
```{r}
baseline_ucb_results_df <- data.frame(
  time = seq_along(baseline_ucb_results$cumulative_reward),
  cumulative_reward = baseline_ucb_results$cumulative_reward,
  cumulative_regret = baseline_ucb_results$cumulative_regret,
  model = "Baseline UCB"
)
```

### Adavanced Model
```{r}
system.time({
  res_adv_ucb_common <- ucb_advanced(
    K, N, mu, y_global, z_global,
    batch_size = 100, burn = 1000, n_iter = 200,
    dynamics = "common"
  )
})
```


Store in df:
```{r}
adv_ucb_results_df <- data.frame(
  time = seq_along(res_adv_ucb_common$cumulative_reward),
  cumulative_reward = res_adv_ucb_common$cumulative_reward,
  cumulative_regret = res_adv_ucb_common$cumulative_regret,
  model = "Advanced UCB"
)
```

Combining results into one df:
```{r}
ucb_compare_df_s1 <- bind_rows(baseline_ucb_results_df, adv_ucb_results_df)
```

### Cumulative Reward:
```{r}
plotucb<-ggplot(ucb_compare_df_s1, aes(x = time, y = cumulative_reward, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Reward: Baseline vs Advanced UCB",
       x = "Time",
       y = "Cumulative Reward") +
  theme_minimal()

ggsave(file.path(output_dir, "ucb_cumulative_reward_single.png"), plot = plotucb, width = 8, height = 6, dpi = 300)
```

### Cumulative Regret:
```{r}
plotucb<-ggplot(ucb_compare_df_s1, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs Advanced UCB",
       x = "Time",
       y = "Cumulative Regret") +
  theme_minimal()

ggsave(file.path(output_dir, "ucb_cumulative_regret_single.png"), plot = plotucb, width = 8, height = 6, dpi = 300)
```

```{r}
ucb_compare_df_s1 <- ucb_compare_df_s1 %>%
  group_by(model) %>%
  arrange(time) %>%
  mutate(inst_regret= c(0, diff(cumulative_regret)))

ggplot(ucb_compare_df_s1, aes(x = time, y = inst_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Instantaneous Regret: Baseline vs Advanced UCB",
       x = "Time",
       y = "Instantaneous Regret") +
  theme_minimal()
```


## 2.3 KFAS AR

### 2.3.1 TS KFAS

#### Baseline

```{r}
baseline_ts_results <- bandit_baselines("ts", K,N,y_global,z_global,mu,
                                        dynamics = "common", batch_size = 1)
baseline_ts_results_df <- data.frame(
  time               = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward  = baseline_ts_results$cumulative_reward,
  cumulative_regret  = baseline_ts_results$cumulative_regret,
  model              = "Baseline TS")
```

### AR

```{r}
res_kfas_ar <- thompson_ar_kfas(
  K,N,mu,y_global,z_global,
  batch_size = 100, burn = 1000, n_iter = 200,
  dynamics   = "common")
kfas_ar_results_df <- data.frame(
  time               = seq_along(res_kfas_ar$cumulative_reward),
  cumulative_reward  = res_kfas_ar$cumulative_reward,
  cumulative_regret  = res_kfas_ar$cumulative_regret,
  model              = "KFAS AR TS")
```


```{r}
ts_compare_df <- bind_rows(baseline_ts_results_df, kfas_ar_results_df)

ggplot(ts_compare_df, aes(time, cumulative_reward, colour = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Reward: Baseline vs KFAS AR(1) TS",
       y = "Cumulative Reward") +
  theme_minimal()

ggplot(ts_compare_df, aes(time, cumulative_regret, colour = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs KFAS AR(1) TS",
       y = "Cumulative Regret") +
  theme_minimal()
```
### 2.3.2 UCB KFAS

```{r}
res_kfas_ucb <- ucb_ar_kfas(
  K,N,mu,y_global,z_global,
  batch_size = 100, burn = 1000, n_iter = 200,
  dynamics   = "common")
kfas_ucb_results_df <- data.frame(
  time               = seq_along(res_kfas_ucb$cumulative_reward),
  cumulative_reward  = res_kfas_ucb$cumulative_reward,
  cumulative_regret  = res_kfas_ucb$cumulative_regret,
  model              = "KFAS AR UCB")
```


```{r}
ucb_compare_df <- bind_rows(baseline_ucb_results_df, kfas_ucb_results_df)

ggplot(ucb_compare_df, aes(time, cumulative_reward, colour = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Reward: Baseline vs KFAS AR(1) UCB",
       y = "Cumulative Reward") +
  theme_minimal()

ggplot(ucb_compare_df, aes(time, cumulative_regret, colour = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs KFAS AR(1) UCB",
       y = "Cumulative Regret") +
  theme_minimal()
```


# 3. Multiple Run - Average Model Comparison

## 3.1 Generate 50 datasets (only if not already generated)!

```{r}
n_runs <- 30
generate_global_datasets(K = K, N = N, mu = mu, pi_global = pi_global, scenario_name = "A", n_runs = n_runs)
```

## 3.2 TS

Run simulations in parallel:
```{r}
ts_runs_df <- future_lapply(seq_len(n_runs), function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "baseline",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
      simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "poor",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "advanced",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
    simulate_model_on_run(
      run_id=i, N=N, K=K,
      algorithm="ts", complexity="kfas",  
      dynamics="common",
      data_path="data_global/A", setting="global")
  )
}, future.seed = TRUE) |> bind_rows()
```

Combine:

```{r}
ts_summary <- ts_runs_df |>
  group_by(time, model_id) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  ) |>
  arrange(model_id, time) |>
  group_by(model_id) |>
  mutate(
    inst_regret = avg_regret - lag(avg_regret),
    inst_regret = if_else(is.na(inst_regret), 0, inst_regret)
  )

```


Comparison:
```{r}
plots<-plot_cumulative_regret(ts_summary) 

ggsave(file.path(output_dir, "ts_cumulative_regret.png"), plot = plots$cumulative, width = 8, height = 6, dpi = 300)
ggsave(file.path(output_dir, "ts_instantaneous_regret.png"), plot = plots$instantaneous, width = 8, height = 6, dpi = 300)
```


## 3.3 UCB

Run simulations in parallel:
```{r}
ucb_runs_df <- future_lapply(seq_len(n_runs), function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "baseline",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "poor",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "advanced",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
    simulate_model_on_run(
      run_id=i, N=N, K=K,
      algorithm="ucb", complexity="kfas",
      dynamics="common",
      data_path="data_global/A", setting="global")
  )
}, future.seed = TRUE) |> bind_rows()
```

Combine:
```{r}
ucb_summary <- ucb_runs_df |>
  group_by(time, model_id) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  ) |>
  arrange(model_id, time) |>
  group_by(model_id) |>
  mutate(
    inst_regret = avg_regret - lag(avg_regret),
    inst_regret = if_else(is.na(inst_regret), 0, inst_regret)
  )
```

Comparison:
```{r}
plots<-plot_cumulative_regret(ucb_summary) 

ggsave(file.path(output_dir, "ucb_cumulative_regret.png"), plot = plots$cumulative, width = 8, height = 6, dpi = 300)
ggsave(file.path(output_dir, "ucb_instantaneous_regret.png"), plot = plots$instantaneous, width = 8, height = 6, dpi = 300)
```

```{r}
ts_runs_df <- simulate_model_on_run(
  run_id     = 1,
  N          = N,
  K          = K,
  algorithm  = "ts",
  complexity = "kfas",
  dynamics   = "common",
  data_path  = "data_global/A",
  setting    = "global"
)
```


```{r}
ts_runs_df_ar <- future_lapply(seq_len(n_runs), function(i) {

  simulate_model_on_run(
    run_id     = i,
    N          = N,
    K          = K,
    algorithm  = "ts",
    complexity = "kfas",   
    dynamics   = "common",
    data_path  = "data_global/A",
    setting    = "global"
  )

}, future.seed = TRUE) |> 
  bind_rows()
```


```{r}
ts_summary_ar <- ts_runs_df_ar |>
  group_by(time, model_id) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  ) |>
  arrange(model_id, time) |>
  group_by(model_id) |>
  mutate(
    inst_regret = avg_regret - lag(avg_regret),
    inst_regret = if_else(is.na(inst_regret), 0, inst_regret)
  )
```

```{r}
plot_cumulative_regret(ts_summary_ar) 
```


