---
title: "Global State Bandit Simulation – KFAS AR(1) vs. Baseline TS"
author: "Marvin Ernst, Oriol Gelabert, Melisa Vadenja"
date: "`r Sys.Date()`"
output: html_document
---

only once:
```{r}
#install.packages("KFAS")
```


```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
library(coda)
library(tidyverse)
library(here)
library(glue)
library(future.apply)
library(stats)
library(KFAS)
library(nimble)                  
plan(multisession)
```

```{r}
source(here("src/load_all.R"))
```

Setting a seed:
```{r}
set.seed(123) 
```

Number of arms:
```{r}
K <- 2
```

Number of time steps:
```{r}
N <- 5000
```

# LOCAL SETTING 

```{r}
pi_list <- list(
  matrix(c(0.997, 0.003,
           0.003, 0.997), nrow = 2, byrow = TRUE),  
  matrix(c(0.997, 0.003,
           0.003, 0.997), nrow = 2, byrow = TRUE)  
)
```

```{r}
mu <- matrix(c(
  0.95, 0.1,  #
  0.5,  0.7   
), nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Local States

Generate and save a single local dataset:

```{r}
args(generate_local_datasets)
generate_local_datasets(K = K, N = 5000, mu = mu, pi_list = pi_list, n_runs = 1, scenario_name = "single_run", root_path = "data_local")
```

Load the generated data:
```{r}
truth <- readRDS("data_local/single_run/local_truth_1.rds")
y_local <- truth$y
z_local <- truth$z
```
## 1.1 Visualize Latent States

```{r visualize-local-states}
z_df <- as.data.frame(z_local)
z_df <- z_df |> mutate(arm = factor(1:K)) |>
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "state") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(z_df, aes(x = time, y = state, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Latent Local States Over Time",
       x = "Time", y = "State")
```

## 1.2 Visualize Rewards
```{r visualize-local-rewards}
y_df <- as.data.frame(y_local)
y_df <- y_df |> mutate(arm = factor(1:K)) |>
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "reward") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(y_df, aes(x = time, y = reward, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Reward Streams Per Arm",
       x = "Time", y = "Reward")
```

# 2. Single Run - Model Comparison

## 2.1 Thompson Sampling

### Baseline

```{r}
baseline_ts_results <- bandit_baselines("ts", K, N, y_local, z_local, mu, dynamics = "independent", batch_size = 100)
```

Store in df:
```{r}
baseline_ts_df <- data.frame(
  time = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward = baseline_ts_results$cumulative_reward,
  cumulative_regret = baseline_ts_results$cumulative_regret,
  model = "Baseline TS"
)
```


## 2.2 KFAS AR Thompson Sampling:


```{r}
res_ts_ar <- thompson_ar_kfas(K, N, mu, y_local, z_local, batch_size = 100, burn = 500, n_iter = 100, dynamics = "independent")
```

```{r}
ts_ar_df <- data.frame(
  time = seq_along(res_poor_ts$cumulative_reward),
  cumulative_reward = res_ts_ar$cumulative_reward,
  cumulative_regret = res_ts_ar$cumulative_regret,
  model = "AR TS"
)
ts_ar_compare_df <- bind_rows(baseline_ts_df, ts_ar_df)
```

```{r}
ggplot(ts_ar_compare_df, aes(x = time, y = cumulative_reward, color = model)) +
  geom_line() +
  labs(title = "Cumulative Reward: Baseline vs KFAS AR TS",
       x = "Time", y = "Cumulative Reward") +
  theme_minimal()
```

```{r}
ggplot(ts_ar_compare_df, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line() +
  labs(title = "Cumulative Regret: Baseline vs KFAS AR TS",
       x = "Time", y = "Cumulative Regret") +
  theme_minimal()
```
## 2.3 UCB

### Baseline

```{r}
baseline_ucb_results <- bandit_baselines("ucb-tuned", K, N, y_local, z_local, mu,
                                         dynamics = "independent", batch_size = 100)
```

Store in df:
```{r}
baseline_ucb_df <- data.frame(
  time = seq_along(baseline_ucb_results$cumulative_reward),
  cumulative_reward = baseline_ucb_results$cumulative_reward,
  cumulative_regret = baseline_ucb_results$cumulative_regret,
  model = "Baseline UCB"
)
```

## 2.4 KFAS AR UCB

```{r}
res_ar_ucb <- ucb_ar_kfas(K, N, mu, y_local, z_local, batch_size = 100, burn = 500, n_iter = 100, dynamics = "independent")
```

```{r}
ar_ucb_df <- data.frame(
  time = seq_along(res_ar_ucb$cumulative_reward),
  cumulative_reward = res_ar_ucb$cumulative_reward,
  cumulative_regret = res_ar_ucb$cumulative_regret,
  model = "AR UCB"
)
ucb_ar_compare_df <- bind_rows(baseline_ucb_df, ar_ucb_df)

```

```{r}
ggplot(ucb_ar_compare_df, aes(x = time, y = cumulative_reward, color = model)) +
  geom_line() +
  labs(title = "Cumulative Reward: Baseline vs KFAS AR UCB",
       x = "Time", y = "Cumulative Reward") +
  theme_minimal()
```
```{r}
ggplot(ucb_ar_compare_df, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line() +
  labs(title = "Cumulative Regret: Baseline vs AR UCB",
       x = "Time", y = "Cumulative Regret") +
  theme_minimal()
```

# 3. Multiple Run - Averaged Performance 

## 3.1 Generate Datasets
```{r}
generate_local_datasets(K = K, N = N, mu = mu, pi_list = pi_list,
                        n_runs = 25,
                        scenario_name = "A",
                        root_path = "data_local")
```

### KFAS TS

```{r}
library(nimble)
ar_multi_df <- future_lapply(1:25, function(i) {
  simulate_model_on_run(
    run_id = i, N = N, K = K,
    algorithm = "ts", complexity = "kfas",
    dynamics = "independent",
    setting = "local",
    data_path = "data_local/A"
  )
}) |> bind_rows()
```

## 3.3 Summary Plot
```{r}
ar_summary <- ts_multi_df |> group_by(time, model) |> summarise(
  avg_regret = mean(cumulative_regret),
  avg_reward = mean(cumulative_reward),
  .groups = "drop"
)

ggplot(ts_summary, aes(x = time, y = avg_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Average Cumulative Regret over 25 Runs (Local State)",
       x = "Time", y = "Cumulative Regret") +
  theme_minimal()
```

### KFAS UCB

```{r}
system.time(
  simulate_model_on_run(
    run_id     = 1,
    algorithm  = "ts",
    complexity = "kfas",
    dynamics   = "independent",
    setting    = "local",
    data_path  = "data_local/A"
  )
)
```


```{r}
ar_ucb_multi_df <- future_lapply(1:25, function(i) {
  simulate_model_on_run(
    run_id = i, N = N, K = K,
    algorithm = "ucb", complexity = "kfas",
    dynamics = "independent",
    setting = "local",
    data_path = "data_local/A"
  )
}) |> bind_rows()
```
```{r}
ar_summary <- ar_ucb_multi_df |> group_by(time, model) |> summarise(
  avg_regret = mean(cumulative_regret),
  avg_reward = mean(cumulative_reward),
  .groups = "drop"
)

ggplot(ts_summary, aes(x = time, y = avg_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Average Cumulative Regret over 25 Runs (Local State)",
       x = "Time", y = "Cumulative Regret") +
  theme_minimal()
```


# Problem Setup: Global Latent State

We simulate a $K$-armed bandit problem where **all arms share the same underlying latent state process** governed by a Hidden Markov Model (HMM). This represents a scenario where there's a common environment or external condition that affects the reward probability of all arms.

### Loading all the Models and Functions


---


Transition Probability:
```{r}
pi_global <- matrix(c(0.999, 0.001,
                     0.001, 0.999),
                     nrow = 2, byrow = TRUE)  
```

Probability of Reward:
```{r}
mu <- matrix(c(0.95, 0.1,
               0.1, 0.95),
                nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Global State

## 1.1 Simulate Ground Truth

For only one Run:
```{r simulate-global-state}
generate_global_datasets(
  K = 2,
  N = N,
  mu = mu,
  pi_global = pi_global,
  n_runs = 1,
  scenario_name = "single_run",
  root_path = "data_global"
)
```

Load the generated data:
```{r}
truth <- readRDS("data_global/single_run/global_truth_1.rds")
y_global <- truth$y
z_global <- truth$z
```


# 1.2 Visualize Latent States

```{r visualize-state}
qplot(1:N, z_global, geom = "line") +
  labs(title = "Latent Global State Over Time",
       x = "Time", y = "State")
```

## 1.3 Visualize Rewards Per Arm

```{r visualize-rewards}
y_df <- as.data.frame(y_global)
y_df <- y_df |> mutate(arm = factor(1:K)) |> 
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "reward") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(y_df, aes(x = time, y = reward, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Reward Streams Per Arm",
       x = "Time", y = "Reward")
```

---

# 2. Single Run - Model Comparison

## 2.1 Thompson Sampling

### Baseline
```{r}
baseline_ts_results <- bandit_baselines("ts", 
                                        K, N, y_global, z_global, mu, 
                                        dynamics = "common",
                                        batch_size = 1)
```

Store in df:
```{r}
baseline_ts_results_df <- data.frame(
  time = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward = baseline_ts_results$cumulative_reward,
  cumulative_regret = baseline_ts_results$cumulative_regret,
  model = "Baseline TS"
)
```

### Adavanced Model
```{r}
system.time({
res_kfas_ar <- thompson_ar_kfas(K, N, mu, y_global, z_global, batch_size = 10)
})
```

Store in df:
```{r}
kfas_ar_results_df <- data.frame(
  time = seq_along(res_kfas_ar$cumulative_reward),
  cumulative_reward = res_kfas_ar$cumulative_reward,
  cumulative_regret = res_kfas_ar$cumulative_regret,
  model = "KFAS AR(1) TS"
)
```

Combining results into one df:
```{r}
ts_compare_df <- bind_rows(baseline_ts_results_df, kfas_ar_results_df)
```

### Cumulative Reward:
```{r}
ggplot(ts_compare_df, aes(x = time, y = cumulative_reward, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Reward: Baseline vs KFAS AR(1) Thompson Sampling",
       x = "Time",
       y = "Cumulative Reward") +
  theme_minimal()
```

### Cumulative Regret:
```{r}
ggplot(ts_compare_df, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs KFAS AR(1) Thompson Sampling",
       x = "Time",
       y = "Cumulative Regret") +
  theme_minimal()
```


## Several Runs 

## 3.1 Generate Datasets
```{r}
generate_global_datasets(K = K, N = N, mu = mu, pi_global = pi_global,
                        n_runs = 25,
                        scenario_name = "A",
                        root_path = "data_global")
```

### Baseline

```{r}
baseline_ts_results <- bandit_baselines("ts", K, N, y_global, z_global, mu, dynamics = "common", batch_size = 100)
```

Store in df:
```{r}
baseline_ts_df <- data.frame(
  time = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward = baseline_ts_results$cumulative_reward,
  cumulative_regret = baseline_ts_results$cumulative_regret,
  model = "Baseline TS"
)
```


```{r}
ar_multi_df <- future_lapply(1:25, function(i) {
  simulate_model_on_run(
    run_id = i, N = N, K = K,
    algorithm = "ts", complexity = "kfas",
    dynamics = "common",
    setting = "global",
    data_path = "data_global/A"
  )
}) |> bind_rows()
```
```{r}
ts_compare_df_s1 <- bind_rows(baseline_ts_df, ar_multi_df)
```

```{r}

ggplot(ts_compare_df_s1, aes(x = time, y = cumulative_regret, color = model)) +
  geom_line(size = 1) +
  labs(title = "Cumulative Regret: Baseline vs KFAS AR(1) Thompson Sampling",
       x = "Time",
       y = "Cumulative Regret") +
  theme_minimal()
```

```{r}
ts_runs_df <- future_lapply(seq_len(n_runs), function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = 2,
      algorithm = "ts",
      complexity = "baseline",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
      simulate_model_on_run(
      run_id = i, N = N, K = 2,
      algorithm = "ts",
      complexity = "kfas",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    )
  )
}, future.seed = TRUE) |> bind_rows()
```
```{r}
ts_summary
```

```{r}
ts_summary <- ts_runs_df |>  group_by(time, model) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  )

```

Comparison:
```{r}
plot_df <- ts_summary |>
  dplyr::rename(model_id = model)  

plot_cumulative_regret(plot_df, title = "Average regret")
```


## UCB

```{r}
ucb_runs_df <- future_lapply(seq_len(n_runs), function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = 2,
      algorithm = "ucb",
      complexity = "baseline",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    ),
      simulate_model_on_run(
      run_id = i, N = N, K = 2,
      algorithm = "ucb",
      complexity = "kfas",
      dynamics = "common",
      data_path = "data_global/A",
      setting = "global"
    )
  )
}, future.seed = TRUE) |> bind_rows()
```

```{r}
ucb_summary <- ucb_runs_df |>  group_by(time, model) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  )

```

Comparison:
```{r}
plot_df <- ucb_summary |>
  dplyr::rename(model_id = model)  

plot_cumulative_regret(plot_df, title = "Average regret")
```