---
title: "Local State Bandit Simulation"
author: "Marvin Ernst, Oriol Gelabert, Melisa Vadenja"
date: "`r Sys.Date()`"
output: html\_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nimble)
library(rjags)
library(coda)
library(tidyverse)
library(here)
library(future.apply)
plan(multisession)
```

```{r}
output_dir <- "local_plots"

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}
```


# Problem Setup: Local Latent State

We simulate a \$K\$-armed bandit problem where **each arm has its own independent latent state process**, governed by separate Hidden Markov Models (HMMs). This setup reflects settings where arms evolve independently in time.

### Loading all the Models and Functions

```{r}
source(here("src/load_all.R"))
```

Setting a seed:
```{r}
set.seed(123)
```

---

# LOCAL SETTING

Number of arms:

```{r}
K <- 2
```

Number of time steps:

```{r}
N <- 5000
```

Transition Probabilities (one per arm):

```{r}
pi_list <- list(
  matrix(c(0.999, 0.001,
           0.001, 0.999), nrow = 2, byrow = TRUE),  
  matrix(c(0.999, 0.001,
           0.001, 0.999),, nrow = 2, byrow = TRUE)  
)
```

Reward Probabilities:

```{r}
mu <- matrix(c(0.95, 0.1,
               0.1, 0.95), nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Local States


## 1.1 Simulate Ground Truth

Generate and save a single local dataset:

```{r}
args(generate_local_datasets)
generate_local_datasets(K = K, N = 5000, mu = mu, pi_list = pi_list, n_runs = 1, scenario_name = "single_run", root_path = "data_local")
```

Load the generated data:
```{r}
truth <- readRDS("data_local/single_run/local_truth_1.rds")
y_local <- truth$y
z_local <- truth$z
```

## 1.2 Visualize Latent States

```{r visualize-local-states}
z_df <- as.data.frame(z_local)
z_df <- z_df |> mutate(arm = factor(1:K)) |>
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "state") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(z_df, aes(x = time, y = state, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Latent Local States Over Time",
       x = "Time", y = "State")
```

## 1.3 Visualize Rewards per Arm
```{r visualize-local-rewards}
y_df <- as.data.frame(y_local)
y_df <- y_df |> mutate(arm = factor(1:K)) |>
  pivot_longer(cols = starts_with("V"), names_to = "time", values_to = "reward") |>
  mutate(time = as.numeric(gsub("V", "", time)))

ggplot(y_df, aes(x = time, y = reward, color = arm)) +
  geom_line() +
  facet_wrap(~ arm, ncol = 1) +
  labs(title = "Reward Streams Per Arm",
       x = "Time", y = "Reward")
```

---

# 2. Single Run - Model Comparison

## 2.1 Thompson Sampling

### Baseline

```{r}
baseline_ts_results <- bandit_baselines("ts", K, N, y_local, z_local, mu, dynamics = "independent", batch_size = 100)
```

Store in df:
```{r}
baseline_ts_df <- data.frame(
  time = seq_along(baseline_ts_results$cumulative_reward),
  cumulative_reward = baseline_ts_results$cumulative_reward,
  cumulative_regret = baseline_ts_results$cumulative_regret,
  model_id = "M0 TS"
)
```

### Poor

```{r}

res_poor_ts <- thompson_poor(K, N, mu, y_local, z_local, batch_size = 100, burn = 500, n_iter = 100, dynamics = "independent")
```

Store in df:
```{r}
poor_ts_df <- data.frame(
  time = seq_along(res_poor_ts$cumulative_reward),
  cumulative_reward = res_poor_ts$cumulative_reward,
  cumulative_regret = res_poor_ts$cumulative_regret,
  model_id = "M1 TS"
)
```

Combining results into one df:
```{r}
ts_compare_df <- bind_rows(baseline_ts_df, poor_ts_df)
```

### Cumulative Reward:
```{r}
plotts<-plot_cumulative_regret(ts_compare_df)
ggsave(file.path(output_dir, "poor_ts_cumulative_regret_single.png"), plot = plotts, width = 8, height = 6, dpi = 300)
```


## 2.2 AR Thompson Sampling:


```{r}
res_ts_ar <- thompson_ar_kfas(K, N, mu, y_local, z_local, batch_size = 100, burn = 500, n_iter = 100, dynamics = "independent")
```

```{r}
ts_ar_df <- data.frame(
  time = seq_along(res_poor_ts$cumulative_reward),
  cumulative_reward = res_ts_ar$cumulative_reward,
  cumulative_regret = res_ts_ar$cumulative_regret,
  model_id = "AR TS"
)
ts_ar_compare_df <- bind_rows(baseline_ts_df, ts_ar_df)
```

```{r}
ts_ar_compare_df
```


```{r}

plotucb<-plot_cumulative_regret(ts_ar_compare_df) 
ggsave(file.path(output_dir, "ar_ts_cumulative_regret_single.png"), plot = plotucb, width = 8, height = 6, dpi = 300)
```

## 2.3 UCB

### Baseline

```{r}
baseline_ucb_results <- bandit_baselines("ucb-tuned", K, N, y_local, z_local, mu,
                                         dynamics = "independent", batch_size = 100)
```

Store in df:
```{r}
baseline_ucb_df <- data.frame(
  time = seq_along(baseline_ucb_results$cumulative_reward),
  cumulative_reward = baseline_ucb_results$cumulative_reward,
  cumulative_regret = baseline_ucb_results$cumulative_regret,
  model_id = "M0 UCB"
)
```

### Poor UCB
```{r}
res_poor_ucb <- ucb_poor(K, N, mu, y_local, z_local, batch_size = 100,
                            burn = 500, n_iter = 100, dynamics = "independent")
```

Store in df:
```{r}
poor_ucb_df <- data.frame(
  time = seq_along(res_poor_ucb$cumulative_reward),
  cumulative_reward = res_poor_ucb$cumulative_reward,
  cumulative_regret = res_poor_ucb$cumulative_regret,
  model_id = "M1 UCB"
)
```

Combining results into one df:
```{r}
ucb_compare_df <- bind_rows(baseline_ucb_df, poor_ucb_df)
```

### Cumulative Regret:
```{r}
plotucb<-plot_cumulative_regret(ucb_compare_df)
ggsave(file.path(output_dir, "poor_ucb_cumulative_regret_single.png"), plot = plotucb, width = 8, height = 6, dpi = 300)
```

```{r}
ucb_compare_df <- ucb_compare_df %>%
  group_by(model_id) %>%
  arrange(time) %>%
  mutate(inst_regret= c(0, diff(cumulative_regret)))

ggplot(ucb_compare_df, aes(x = time, y = inst_regret, color = model_id)) +
  geom_line(size = 1) +
  labs(title = "Instantaneous Regret: Baseline vs Advanced UCB",
       x = "Time",
       y = "Instantaneous Regret") +
  theme_minimal()
```


## 2.4 AR UCB

```{r}
res_ar_ucb <- ucb_ar_kfas(K, N, mu, y_local, z_local, batch_size = 100, burn = 500, n_iter = 100, dynamics = "independent")
```

```{r}
ar_ucb_df <- data.frame(
  time = seq_along(res_ar_ucb$cumulative_reward),
  cumulative_reward = res_ar_ucb$cumulative_reward,
  cumulative_regret = res_ar_ucb$cumulative_regret,
  model_id = "AR UCB"
)
ucb_ar_compare_df <- bind_rows(baseline_ucb_df, ar_ucb_df)

```


```{r}
plt<- plot_cumulative_regret(ucb_ar_compare_df)
ggsave(file.path(output_dir, "ar_ucb_cumulative_regret_single.png"), plot = plt, width = 8, height = 6, dpi = 300)
```


# 3. Multiple Run - Averaged Performance 

#### DONT FORGET TO CHANGE NR OF RUNS IN PARALLELIZATION FUNCTION

## 3.1 Generate Datasets (for 30 runs)
```{r}
generate_local_datasets(K = K, N = N, mu = mu, pi_list = pi_list,
                        n_runs = 30,
                        scenario_name = "A",
                        root_path = "data_local")
```

## 3.2 Parallel Runs

### 3.2.1 TS


## issues cuz function is not defined like this.

```{r}
ts_runs_df <- future_lapply(1:30, function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "baseline",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
      simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "poor",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ts",
      complexity = "advanced",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
    simulate_model_on_run(
      run_id=i, N=N, K=K,
      algorithm="ts", complexity="kfas",  
      dynamics="independent",
      data_path="data_local/A", setting="local")
  )
}, future.seed = TRUE) |> bind_rows()
```



## 3.3 Summary Plot

```{r}
ts_summary <- ts_runs_df |>
  group_by(time, model_id) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  ) |>
  arrange(model_id, time) |>
  group_by(model_id) |>
  mutate(
    inst_regret = avg_regret - lag(avg_regret),
    inst_regret = if_else(is.na(inst_regret), 0, inst_regret)
  )

ggplot(ts_summary, aes(x = time, y = avg_regret, color = model_id)) +
  geom_line(size = 1) +
  labs(title = "Average Cumulative Regret over 25 Runs (Local State)",
       x = "Time", y = "Cumulative Regret") +
  theme_minimal()

```
```{r}
plots<-plot_cumulative_regret(ts_summary) 

ggsave(file.path(output_dir, "ts_runs_cumulative_regret.png"), plot = plots$cumulative, width = 8, height = 6, dpi = 300)
ggsave(file.path(output_dir, "ts_runs_instantaneous_regret.png"), plot = plots$instantaneous, width = 8, height = 6, dpi = 300)
```

### 3.2.2 UCB
```{r}
ucb_runs_df <- future_lapply(1:30, function(i) {
  bind_rows(
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "baseline",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
      simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "poor",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
    simulate_model_on_run(
      run_id = i, N = N, K = K,
      algorithm = "ucb",
      complexity = "advanced",
      dynamics = "independent",
      data_path = "data_local/A",
      setting = "local"
    ),
    simulate_model_on_run(
      run_id=i, N=N, K=K,
      algorithm="ucb", complexity="kfas",  
      dynamics="independent",
      data_path="data_local/A", setting="local")
  )
}, future.seed = TRUE) |> bind_rows()
```
```{r}
ucb_summary <- ucb_runs_df |>
  group_by(time, model_id) |>
  summarise(
    avg_regret = mean(cumulative_regret),
    avg_reward = mean(cumulative_reward),
    .groups = "drop"
  ) |>
  arrange(model_id, time) |>
  group_by(model_id) |>
  mutate(
    inst_regret = avg_regret - lag(avg_regret),
    inst_regret = if_else(is.na(inst_regret), 0, inst_regret)
  )
```


```{r}
plot_cumulative_regret(ucb_summary) 

plots<-plot_cumulative_regret(ucb_summary) 

ggsave(file.path(output_dir, "ucb_runs_cumulative_regret.png"), plot = plots$cumulative, width = 8, height = 6, dpi = 300)
ggsave(file.path(output_dir, "ucb_runs_instantaneous_regret.png"), plot = plots$instantaneous, width = 8, height = 6, dpi = 300)
```
