---
title: "a"
output: html_document
date: "2025-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
library(coda)
library(tidyverse)
library(here)
library(nimble)
library(glue)
library(future.apply)
plan(multisession)
```


# Problem Setup: Global Latent State

We simulate a $K$-armed bandit problem where **all arms share the same underlying latent state process** governed by a Hidden Markov Model (HMM). This represents a scenario where there's a common environment or external condition that affects the reward probability of all arms.

### Loading all the Models and Functions
```{r}
source(here("src/load_all.R"))
```

Setting a seed:
```{r}
set.seed(123)
```

---

# GLOABL SETTING

Number of arms:
```{r}
K <- 2 
```

Number of time steps:
```{r}
N <- 10000
```

Transition Probability:
```{r}
pi_global <- matrix(c(0.9, 0.1,
                     0.1, 0.9),
                     nrow = 2, byrow = TRUE)  
```

Probability of Reward:
```{r}
mu <- matrix(c(0.1, 0.95,
               0.95, 0.1),
                nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Global State

## 1.1 Simulate Ground Truth

For only one Run:
```{r simulate-global-state}
generate_global_datasets(
  K = K,
  N = N,
  mu = mu,
  pi_global = pi_global,
  n_runs = 1,
  scenario_name = "single_run",
  root_path = "data_global"
)
```

Load the generated data:
```{r}
truth <- readRDS("data_global/single_run/global_truth_1.rds")
y_global <- truth$y
z_global <- truth$z
```

```{r cars}

  model_path <- advanced_model_path
  
  observed_rewards <- list()
  selected_arms <- numeric(N)
  received_rewards <- numeric(N)
  regret <- numeric(N)
  post_matrix <- NULL
  
  z_true <- z_global
  y_true <- y_global
  dynamics<-'common'
  batch_size <- 100
  burn = 500
  n_iter = 100
  
  data_track <-list()
  
  # Oracle computation
  mu_selected <- matrix(0, nrow = K, ncol = N)
  if (dynamics == "common") {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[t] + 1]
      }
    }
  } else {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[i, t] + 1]
      }
    }
  }
  oracle_arm <- apply(mu_selected, 2, which.max)
  
  # Initialization
  selected_arms[1] <- sample(1:K, 1)
  observed_data <- matrix(NA, nrow = K, ncol = N)
  observed_data[selected_arms[1], 1] <- y_true[selected_arms[1], 1]
  received_rewards[1] <- y_true[selected_arms[1], 1]
  
  regret[1] <- mu_selected[oracle_arm[1], 1] - 
    if (dynamics == "common") mu[selected_arms[1], z_true[1] + 1] else mu[selected_arms[1], z_true[selected_arms[1], 1] + 1]
  
  # TS loop
  for (t in 2:N) {
    sampled_values <- numeric(K)
    
    if (sum(!is.na(observed_data)) >= K * 2 && t %% batch_size == 0) {
      data_list <- list(
        y_obs = observed_data[, 1:(t-1)],
        K = K,
        N = t - 1
      )
      
      data_track[[t/batch_size]] <- data_list
      model <- jags.model(model_path, data = data_list, n.chains = 1, quiet = TRUE)
      update(model, burn)
      post <- coda.samples(model, c("mu", "pi", paste0("z[", t - 1, "]")), n.iter = n_iter)
      post_matrix <- as.matrix(post)
      message(glue::glue("[{Sys.time()}] Performed inference at time step {t}"))
    }
    
    if (is.matrix(post_matrix)) {
      idx <- sample(1:nrow(post_matrix), 1)
      pi0 <- post_matrix[idx, "pi[1]"]
      pi1 <- post_matrix[idx, "pi[2]"]
      if (t %% batch_size == 0) {
        z_last <- post_matrix[idx, paste0("z[", t - 1, "]")]
      }
      sampled_pi <- ifelse(z_last == 0, pi0, pi1)
      z_t <- rbinom(1, 1, prob = sampled_pi)
      z_last <- z_t
    }
    
    # Arm selection
    for (i in 1:K) {
      if (is.matrix(post_matrix)) {
        mu0 <- post_matrix[idx, paste0("mu[", i, ",1]")]
        mu1 <- post_matrix[idx, paste0("mu[", i, ",2]")]
        sampled_values[i] <- (1 - z_t) * mu0 + z_t * mu1
      } else {
        sampled_values[i] <- runif(1)
      }
    }
    
    selected_arms[t] <- which.max(sampled_values)
    r_t <- y_true[selected_arms[t], t]
    received_rewards[t] <- r_t
    observed_data[selected_arms[t], t] <- r_t
    
    true_z <- if (dynamics == "common") z_true[t] else z_true[selected_arms[t], t]
    regret[t] <- mu_selected[oracle_arm[t], t] - mu[selected_arms[t], true_z + 1]
  }
```

```{r}
library(rjags)
library(coda)
library(ggmcmc)
library(dplyr)
library(ggplot2)

# Parameters
param_names <- c("mu", "pi")

# Result list for ggmcmc
gg_data_all <- list()

# Loop through data_track list
for (i in seq_along(data_track)) {
  cat("Running model for dataset", i, "\n")
  
  model <- jags.model(model_path, data = data_track[[i]], n.chains = 1, quiet = TRUE)
  update(model, burn)
  
  post <- coda.samples(model, variable.names = param_names, n.iter = n_iter)
  
  # Convert to ggmcmc data frame and tag with dataset id
  gg_data <- ggs(post)
  gg_data$dataset <- paste0("Dataset ", i)
  gg_data_all[[i]] <- gg_data
}

# Combine all datasets
gg_data_combined <- bind_rows(gg_data_all)

# Plot traceplots with faceting by parameter and coloring by dataset
ggplot(gg_data_combined, aes(x = Iteration, y = value, color = dataset)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~Parameter, scales = "free_y") +
  theme_minimal() +
  labs(title = "Traceplots by Parameter Across Datasets",
       y = "Parameter Value", x = "Iteration")

```
```{r}
ggplot(gg_data_all[[i]], aes(x = Iteration, y = value, color = dataset)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~Parameter, scales = "free_y") +
  theme_minimal() +
  labs(title = "Traceplots on last Run",
       y = "Parameter Value", x = "Iteration")
```

```{r cars}

  model_path <- advanced_model_cheat_path
  
  observed_rewards <- list()
  selected_arms <- numeric(N)
  received_rewards <- numeric(N)
  regret <- numeric(N)
  post_matrix <- NULL
  
  z_true <- z_global
  y_true <- y_global
  dynamics<-'common'
  batch_size <- 100
  burn = 500
  n_iter = 100
  
  data_track <-list()
  
  # Oracle computation
  mu_selected <- matrix(0, nrow = K, ncol = N)
  if (dynamics == "common") {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[t] + 1]
      }
    }
  } else {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[i, t] + 1]
      }
    }
  }
  oracle_arm <- apply(mu_selected, 2, which.max)
  
  # Initialization
  selected_arms[1] <- sample(1:K, 1)
  observed_data <- matrix(NA, nrow = K, ncol = N)
  observed_data[selected_arms[1], 1] <- y_true[selected_arms[1], 1]
  received_rewards[1] <- y_true[selected_arms[1], 1]
  
  regret[1] <- mu_selected[oracle_arm[1], 1] - 
    if (dynamics == "common") mu[selected_arms[1], z_true[1] + 1] else mu[selected_arms[1], z_true[selected_arms[1], 1] + 1]
  
  # TS loop
  for (t in 2:N) {
    sampled_values <- numeric(K)
    
    if (sum(!is.na(observed_data)) >= K * 2 && t %% batch_size == 0) {
      data_list <- list(
        y_obs = observed_data[, 1:(t-1)],
        K = K,
        N = t - 1
      )
      
      data_track[[t/batch_size]] <- data_list
      model <- jags.model(model_path, data = data_list, n.chains = 1, quiet = TRUE)
      update(model, burn)
      post <- coda.samples(model, c("mu", "pi", paste0("z[", t - 1, "]")), n.iter = n_iter)
      post_matrix <- as.matrix(post)
      message(glue::glue("[{Sys.time()}] Performed inference at time step {t}"))
    }
    
    if (is.matrix(post_matrix)) {
      idx <- sample(1:nrow(post_matrix), 1)
      pi0 <- post_matrix[idx, "pi[1]"]
      pi1 <- post_matrix[idx, "pi[2]"]
      if (t %% batch_size == 0) {
        z_last <- post_matrix[idx, paste0("z[", t - 1, "]")]
      }
      sampled_pi <- ifelse(z_last == 0, pi0, pi1)
      z_t <- rbinom(1, 1, prob = sampled_pi)
      z_last <- z_t
    }
    
    # Arm selection
    for (i in 1:K) {
      if (is.matrix(post_matrix)) {
        mu0 <- post_matrix[idx, paste0("mu[", i, ",1]")]
        mu1 <- post_matrix[idx, paste0("mu[", i, ",2]")]
        sampled_values[i] <- (1 - z_t) * mu0 + z_t * mu1
      } else {
        sampled_values[i] <- runif(1)
      }
    }
    
    selected_arms[t] <- which.max(sampled_values)
    r_t <- y_true[selected_arms[t], t]
    received_rewards[t] <- r_t
    observed_data[selected_arms[t], t] <- r_t
    
    true_z <- if (dynamics == "common") z_true[t] else z_true[selected_arms[t], t]
    regret[t] <- mu_selected[oracle_arm[t], t] - mu[selected_arms[t], true_z + 1]
  }
```



```{r}
library(rjags)
library(coda)
library(ggmcmc)
library(dplyr)
library(ggplot2)

# Parameters
param_names <- c("mu", "pi")

# Result list for ggmcmc
gg_data_all <- list()

# Loop through data_track list
for (i in seq_along(data_track)) {
  cat("Running model for dataset", i, "\n")
  
  model <- jags.model(model_path, data = data_track[[i]], n.chains = 1, quiet = TRUE)
  update(model, burn)
  
  post <- coda.samples(model, variable.names = param_names, n.iter = n_iter)
  
  # Convert to ggmcmc data frame and tag with dataset id
  gg_data <- ggs(post)
  gg_data$dataset <- paste0("Dataset ", i)
  gg_data_all[[i]] <- gg_data
}

# Combine all datasets
gg_data_combined <- bind_rows(gg_data_all)

# Plot traceplots with faceting by parameter and coloring by dataset
ggplot(gg_data_combined, aes(x = Iteration, y = value, color = dataset)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~Parameter, scales = "free_y") +
  theme_minimal() +
  labs(title = "Traceplots by Parameter Across Datasets",
       y = "Parameter Value", x = "Iteration")

```
