---
title: "a"
output: html_document
date: "2025-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
library(coda)
library(tidyverse)
library(here)
library(nimble)
library(glue)
library(future.apply)
plan(multisession)
```


# Problem Setup: Global Latent State

We simulate a $K$-armed bandit problem where **all arms share the same underlying latent state process** governed by a Hidden Markov Model (HMM). This represents a scenario where there's a common environment or external condition that affects the reward probability of all arms.

### Loading all the Models and Functions
```{r}
source(here("src/load_all.R"))
```

Setting a seed:
```{r}
set.seed(123)
```

---

# GLOABL SETTING

Number of arms:
```{r}
K <- 2 
```

Number of time steps:
```{r}
N <- 10000
```

Transition Probability:
```{r}
pi_global <- matrix(c(0.999, 0.001,
                     0.001, 0.999),
                     nrow = 2, byrow = TRUE)  
```

Probability of Reward:
```{r}
mu <- matrix(c(0.95, 0.1,
               0.1, 0.95),
                nrow = K, byrow = TRUE)
```

---

# 1. Generating Ground Truth Data for Global State

## 1.1 Simulate Ground Truth

For only one Run:
```{r simulate-global-state}
generate_global_datasets(
  K = K,
  N = N,
  mu = mu,
  pi_global = pi_global,
  n_runs = 1,
  scenario_name = "single_run",
  root_path = "data_global"
)
```

Load the generated data:
```{r}
truth <- readRDS("data_global/single_run/global_truth_1.rds")
y_global <- truth$y
z_global <- truth$z
```

```{r cars}

  model_path <- advanced_model_path
  
  observed_rewards <- list()
  selected_arms <- numeric(N)
  received_rewards <- numeric(N)
  regret <- numeric(N)
  post_matrix <- NULL
  
  z_true <- z_global
  y_true <- y_global
  dynamics<-'common'
  batch_size <- 100
  burn = 1000
  n_iter = 200
  
  data_track <-list()
  
  # Oracle computation
  mu_selected <- matrix(0, nrow = K, ncol = N)
  if (dynamics == "common") {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[t] + 1]
      }
    }
  } else {
    for (i in 1:K) {
      for (t in 1:N) {
        mu_selected[i, t] <- mu[i, z_true[i, t] + 1]
      }
    }
  }
  oracle_arm <- apply(mu_selected, 2, which.max)
  
  # Initialization
  selected_arms[1] <- sample(1:K, 1)
  observed_data <- matrix(NA, nrow = K, ncol = N)
  observed_data[selected_arms[1], 1] <- y_true[selected_arms[1], 1]
  received_rewards[1] <- y_true[selected_arms[1], 1]
  
  regret[1] <- mu_selected[oracle_arm[1], 1] - 
    if (dynamics == "common") mu[selected_arms[1], z_true[1] + 1] else mu[selected_arms[1], z_true[selected_arms[1], 1] + 1]
  
  # TS loop
  for (t in 2:N) {
    sampled_values <- numeric(K)
    
    if (sum(!is.na(observed_data)) >= K * 2 && t %% batch_size == 0) {
      data_list <- list(
        y_obs = observed_data[, 1:(t-1)],
        K = K,
        N = t - 1
      )
      
      data_track[[t/batch_size]] <- data_list
      model <- jags.model(model_path, data = data_list, n.chains = 1, quiet = TRUE)
      update(model, burn)
      post <- coda.samples(model, c("mu", "pi", paste0("z[", t - 1, "]")), n.iter = n_iter)
      post_matrix <- as.matrix(post)
      message(glue::glue("[{Sys.time()}] Performed inference at time step {t}"))
    }
    
    if (is.matrix(post_matrix)) {
      idx <- sample(1:nrow(post_matrix), 1)
      pi0 <- post_matrix[idx, "pi[1]"]
      pi1 <- post_matrix[idx, "pi[2]"]
      if (t %% batch_size == 0) {
        z_last <- post_matrix[idx, paste0("z[", t - 1, "]")]
      }
      sampled_pi <- ifelse(z_last == 0, pi0, pi1)
      z_t <- rbinom(1, 1, prob = sampled_pi)
      z_last <- z_t
    }
    
    # Arm selection
    for (i in 1:K) {
      if (is.matrix(post_matrix)) {
        mu0 <- post_matrix[idx, paste0("mu[", i, ",1]")]
        mu1 <- post_matrix[idx, paste0("mu[", i, ",2]")]
        sampled_values[i] <- (1 - z_t) * mu0 + z_t * mu1
      } else {
        sampled_values[i] <- runif(1)
      }
    }
    
    selected_arms[t] <- which.max(sampled_values)
    r_t <- y_true[selected_arms[t], t]
    received_rewards[t] <- r_t
    observed_data[selected_arms[t], t] <- r_t
    
    true_z <- if (dynamics == "common") z_true[t] else z_true[selected_arms[t], t]
    regret[t] <- mu_selected[oracle_arm[t], t] - mu[selected_arms[t], true_z + 1]
  }
```
```{r}

means_df <- gg_data_combined %>%
  mutate(run = as.integer(gsub("Dataset ", "", dataset))) %>%
  group_by(run, Parameter) %>%
  summarise(mean_value = mean(value), .groups = "drop")

ggplot(means_df, aes(x = run, y = mean_value)) +
  geom_line(alpha = 0.7, size = 0.9, color = 'black') +
  facet_wrap(~ Parameter, scales = "free_y") +
  labs(
    title = "Posterior Means of Parameters Across Runs",
    x     = "Run (batch index)",
    y     = "Posterior Mean"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none"
  )

#pis here look good but not the mus
```
```{r}
# the issue above was that mus are getting mixed up to i set the lower ones together and the high ones together
tidy_means <- means_df %>%
  filter(str_detect(Parameter, "^mu")) %>%
  extract(Parameter,
          into = c("what","arm","state"),
          regex = "(mu)\\[(\\d+),(\\d+)\\]") %>%
  mutate(across(c(arm, state), as.integer))  

relabeled <- tidy_means %>%
  group_by(run, arm) %>%
  arrange(mean_value) %>%
  mutate(rel_state = row_number()) %>%
  ungroup() %>%
  mutate(newParam = paste0("mu[", arm, ",", rel_state, "]"))

plot_df <- relabeled %>%
  select(run, Parameter = newParam, mean_value)

ggplot(plot_df, aes(x = run, y = mean_value)) +
  geom_line(alpha = 0.7, size = 0.9, color = 'black') +
  facet_wrap(~ Parameter, scales = "free_y") +
  labs(
    title = "Posterior Means of Parameters Across Runs\n(label-switched corrected)",
    x     = "Run (batch index)",
    y     = "Posterior Mean"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```



```{r}
ggplot(gg_data_all[[length(gg_data_all)]], aes(x = Iteration, y = value)) +
  geom_line(alpha = 0.7, size = 0.9, color = 'black') +
  facet_wrap(~Parameter, scales = "free_y") +
  labs(
    title = "Parameter Traceplots after JAGGS run with 10000 samples",
    x = "Sample",
    y = "Parameter Value"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "none"
  )

```
# The Local Setup

Transition Probabilities (one per arm):

```{r}
pi_list <- list(
  matrix(c(0.999, 0.001,
           0.001, 0.999), nrow = 2, byrow = TRUE),  
  matrix(c(0.999, 0.001,
           0.001, 0.999),, nrow = 2, byrow = TRUE)  
)
```

Reward Probabilities:

```{r}
mu <- matrix(c(0.95, 0.1,
               0.1, 0.95), nrow = K, byrow = TRUE)
```

---

## Simulate Ground Truth

Generate and save a single local dataset:

```{r}
generate_local_datasets(K = K, N = N, mu = mu, pi_list = pi_list, n_runs = 1, scenario_name = "single_run", root_path = "data_local")
```

Load the generated data:
```{r}
truth <- readRDS("data_local/single_run/local_truth_1.rds")
y_local <- truth$y
z_local <- truth$z
```

```{r cars}
  N<- 5000

  model_path <- poor_model_path
  
  selected_arms   <- integer(N)
  observed_data   <- matrix(NA, nrow = K, ncol = N)
  received_rewards<- numeric(N)
  regret          <- numeric(N)
  
  # Oracle precompute unchanged
  mu_selected <- matrix(0, nrow = K, ncol = N)
  for (i in 1:K) {
    for (t in 1:N) {
      mu_selected[i,t] <- mu[i, z_true[t] + 1]
    }
  }
  oracle_arm <- apply(mu_selected, 2, which.max)
  
  selected_arms[1]     <- sample(1:K, 1)
  observed_data[,1]    <- NA
  observed_data[selected_arms[1],1] <- y_true[selected_arms[1],1]
  received_rewards[1]  <- y_true[selected_arms[1],1]
  regret[1]            <- mu_selected[oracle_arm[1],1] - 
                           mu[selected_arms[1], z_true[1]+1]
  
  post_matrix <- NULL
  
  # Thompson-sampling loop
  for (t in 2:N) {
    sampled_values <- numeric(K)
  

    if (t %% batch_size == 0) {
      for (i in 1:K) {
        obs_idx <- which(!is.na(observed_data[i, ]))
        if (length(obs_idx) >= 2) {
          data_list <- list(
            y = observed_data[i, obs_idx],
            N = length(obs_idx)
          )

          mdl <- jags.model(
            file     = poor_model_path,
            data     = data_list,
            n.chains = 1,
            quiet    = TRUE
          )
          update(mdl, burn)
          post <- coda.samples(
            mdl,
            variable.names = c("mu0","mu1","pi[1]","pi[2]",
                                paste0("z[", length(obs_idx), "]")),
            n.iter         = n_iter
          )
          post_matrix <- as.matrix(post)

          idx <- sample(nrow(post_matrix), 1)
          mu0 <- post_matrix[idx, "mu0"]
          mu1 <- post_matrix[idx, "mu1"]
          pi_low  <- post_matrix[idx, "pi[1]"]
          pi_high <- post_matrix[idx, "pi[2]"]
          
          z_prev <- ifelse(is.na(observed_data[i, obs_idx[length(obs_idx)]]),
                           0, 
                           observed_data[i, obs_idx[length(obs_idx)]])
          p_z1 <- z_prev*pi_high + (1-z_prev)*pi_low

          sampled_values[i] <- p_z1 * mu1 + (1-p_z1)*mu0
        }
      }
    }
    if (all(sampled_values == 0)) {
      sampled_values <- runif(K)
    }
  

    a_t <- which.max(sampled_values)
    selected_arms[t] <- a_t
  
    r_t <- y_true[a_t, t]
    received_rewards[t] <- r_t
    observed_data[a_t, t] <- r_t
    true_z <- z_true[t]
    regret[t] <- mu_selected[oracle_arm[t], t] - mu[a_t, true_z+1]
  }
  
```
